---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rcamelspe

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/rcamelspe)](https://CRAN.R-project.org/package=rcamelspe)
<!-- badges: end -->

The goal of `rcamelspe` is to provide an ultra-fast, computationally efficient, and user-friendly R interface to download, load, filter, and manage the **CAMELS-PE** (Catchment Attributes and Meteorology for Large-sample Studies - Peru) dataset. 

CAMELS-PE includes daily meteorological and hydrological time series, static catchment attributes, and geospatial representations for **136 catchments** distributed across the Pacific, Atlantic, and Titicaca hydrographic regions of Peru.

To handle large volumes of records (over 2.2 million rows of daily climate observations and simulations) with maximum speed and minimum memory footprint, the package is engineered using:
- **`arrow`**: for memory-mapped dataset reading, predicate pushdown (filtering stations and dates directly at the C++ scan level), and zero-overhead column projection.
- **`collapse`**: for fast C/C++ data frame transformations, grouped calculations, row binding, and relational joins.
- **`RCamelsPE` compatibility**: full drop-in compatibility and aliases (`read_timeseries()`, `read_attributes()`, `read_metadata()`, `read_dictionary()`, `read_geospatial()`, `set_camels_path()`, `get_camels_path()`).

---

## Installation

You can install the released version of `rcamelspe` from CRAN with:

```r
install.packages("rcamelspe")
```

And the development version from GitHub with:

```r
# install.packages("pak")
pak::pak("pefrens/rcamelspe")
```

---

## Bundled Sample Dataset & Instant Exploration

`rcamelspe` includes a lightweight, self-contained sample dataset in `inst/extdata/sample_camels_pe` (~120 KB) covering 2 representative catchments (`PE_110139` and `PE_111151`). This allows you to explore all functions immediately after installation without waiting for large downloads.

When you are ready to work with the complete 136-catchment dataset (approx. 120 MB zip, extracting to 186 MB), you can download it directly from Zenodo (DOI: 10.5281/zenodo.20058779):

```r
library(rcamelspe)

# Download and extract the full dataset to user data directory
download_pe_data()
```

---

## Quick Start Example

```{r example_setup}
library(rcamelspe)
```

### 1. Load Station Metadata & Data Dictionary

Read the gauging station metadata or inspect the unified data dictionary:

```{r load_stations}
# Gauging stations
stations <- load_pe_metadata(type = "stations")
head(stations[, c("gauge_id", "gauge_name", "gauge_region", "gauge_elev")])

# Data dictionary
dict <- read_dictionary(category = "climatic")
head(dict[, c("variable", "description", "unit")])
```

---

### 2. Load and Merge Catchment Attributes

Retrieve topographic, geologic, soil, land cover, climatic indices, hydrological signatures, or human intervention attributes:

```{r load_attributes}
# Load merged topographic and landcover attributes
attrs <- load_pe_attributes(c("topographic", "landcover"))
head(attrs[, c("gauge_id", "area", "elev_mean", "forest_perc")])
```

---

### 3. Load Daily Time Series (Highly Optimized)

Load time series for specific catchments, filter by date range, and select only required variables at scan time:

```{r load_timeseries}
# Load precipitation and observed streamflow with date filter
ts_data <- load_pe_timeseries(
  gauge_ids = "PE_110139", 
  variables = c("prec", "flow_obs"),
  start_date = "2000-01-01",
  end_date = "2000-06-30"
)
head(ts_data)
```

---

### 4. Access Geospatial Layers & Plotting

Load catchment boundaries and station locations as `sf` objects, and plot them:

```{r load_spatial}
# Catchment boundaries
catchments <- load_pe_geospatial(type = "catchments")
gauges <- load_pe_geospatial(type = "gauges")

# Plot time series
p_ts <- plot_pe_timeseries(ts_data, variable = "prec")

# Plot catchments with gauge overlays
p_map <- plot_pe_catchments(catchments, gauges = gauges)
```

---

### 5. Full Drop-in Compatibility with `RCamelsPE`

Existing scripts written for `RCamelsPE` run without any modifications:

```r
# Compatible syntax:
stations <- read_metadata()
dict     <- read_dictionary(category = "topographic")
attrs    <- read_attributes(type = "topographic")
ts       <- read_timeseries(gauge_id = "PE_110139", vars = c("date", "prec", "flow_obs"))
spatial  <- read_geospatial(type = "catchments")
```
