---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rcamelspe

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/rcamelspe)](https://CRAN.R-project.org/package=rcamelspe)
<!-- badges: end -->

The goal of `rcamelspe` is to provide an efficient and user-friendly R interface to download, load, filter, and manage the **CAMELS-PE** (Catchment Attributes and Meteorology for Large-sample Studies - Peru) dataset. 

CAMELS-PE includes daily meteorological and hydrological time series, static catchment attributes, and geospatial representations for **136 catchments** distributed across the Pacific, Atlantic, and Titicaca hydrographic regions of Peru.

To manage the volume of records (over 2.2 million rows of daily climate data), the package is designed using:
- **`arrow`**: for high-performance reading of large files.
- **`collapse`**: for fast data frame subsetting, selection, binding, and join operations.

---

## Installation

You can install the released version of `rcamelspe` from CRAN with:

```r
install.packages("rcamelspe")
```

And the development version from GitHub with:

```r
# install.packages("devtools")
devtools::install_github("pefrens/rcamelspe")
```

---

## Downloading the Dataset

You can programmatically download and extract the dataset (approx. 120 MB zip, extracting to 186 MB) directly from the Zenodo repository (`https://doi.org/10.5281/zenodo.20058779`):

```r
library(rcamelspe)

# Download and extract the dataset to a folder named "data-raw"
download_pe_data(dest_dir = "data-raw")
```

This will automatically download, extract, and register the dataset path in the package session.

---

## Quick Start Example

This example demonstrates how to configure paths, load station metadata, read catchment attributes, load climate daily timeseries, and access geospatial boundaries.

```{r example_setup}
library(rcamelspe)

# If the folder was extracted to a custom location, configure it:
# set_camels_pe_path("data-raw/CAMELS-PE")
```

### 1. Load Station Metadata
Read the list of all 136 gauging stations included in the dataset:

```{r load_stations}
stations <- load_pe_metadata(type = "stations")
head(stations[, c("gauge_id", "gauge_name", "gauge_region", "gauge_elev")])
```

---

### 2. Load and Merge Catchment Attributes
Retrieve topographic, geologic, soil, land cover, climatic indices, hydrological signatures, or human intervention attributes. If `attributes = "all"`, all 7 static attribute tables are merged:

```{r load_attributes}
# Load and merge topographic and landcover attributes
attrs <- load_pe_attributes(c("topographic", "landcover"))
head(attrs[, c("gauge_id", "area", "elev_mean", "forest_perc", "agricul_perc")])
```

---

### 3. Loading daily timeseries (Highly Optimized)
You can load time series for a subset of stations or the entire dataset. 
- **Optimization**: If 10 or fewer catchments are requested, the package reads individual catchment CSV files, avoiding parsing the 186 MB main database. If more are requested, it reads the main database using `arrow` and filters it instantly using `collapse`.

```{r load_timeseries}
# Load precipitation and observed flow for two specific catchments
ts_sub <- load_pe_timeseries(
  gauge_ids = c("PE_110139", "PE_111151"), 
  variables = c("prec", "flow_obs")
)
summary(ts_sub)
```

---

### 4. Access Geospatial Layers
Load catchment boundaries (polygon) or station gauge locations (points) as `sf` spatial objects:

```{r load_spatial}
# Load catchment boundaries for a specific subset of gauges
catchments_sf <- load_pe_geospatial(
  type = "catchments", 
  gauge_ids = c("PE_110139", "PE_111151")
)
print(catchments_sf)
```
